New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-25 paper/paper.tex
some improvements

\begin{document}
\maketitle
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically founded extension to the original estimate is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum-likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle. The new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

Currently, the best algorithm that also works for distributed setups is the near-optimal HyperLogLog algorithm \cite{Flajolet2007} with space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$. The originally proposed estimation method has some problems to guarantee the same estimation error over the entire range of cardinalities. It was proposed to correct the estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.

@@

\section{HyperLogLog data structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers is chosen to be a power of 2: $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$. All registers start with initial value 0. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

@@

For element insertion a hash value with bit size $\symPrecision+\symRegRange$ is required. The consumed number of bits also limits the maximum cardinality that can be tracked. Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy is drastically reduced.

@@

\cref{alg:insert} has some properties which are especially useful in distributed environments. The insertion order of elements has no influence on the final HyperLogLog sketch. Furthermore, HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ can be easily merged. The resulting register values are simply the register-wise maximum values. The final HyperLogLog sketch corresponds to the union of both sets that have been represented by the input sketches.

A $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}. In this case each register value is represented by a single bit. Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).

\begin{algorithm}
\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) which are initialized to 0 and which can represent any integer in range $[0, \symRegRange+1]$.}
\label{alg:insert}
\begin{algorithmic}
\Procedure {InsertElement}{\symDataItem, $\vec{\symRegVal}$}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
\State $\symIndexJ \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
\State $\symRegVal' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
\State $\symRegVal_\symIndexJ \gets\max(\symRegVal_\symIndexJ, \symRegVal')$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
\begin{equation}
\symRegProbability_{\symNu\symKappa} 
:=
\begin{cases}
1 & \symNu=0 \wedge \symKappa = 0\\
0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
0 & \symNu\geq1 \wedge \symKappa = 0\\
\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
\end{cases}
\end{equation}

The order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$. As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation. The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim(\vec{\symCount})-2$, respectively.

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Unfortunately, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze, because the register values are statistically dependent. Therefore, as proposed in \cite{Flajolet2007} a Poisson model can be used, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexJ=1}^\symNumReg
\symRegProbability_
{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
\nonumber\\
&= 
\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left[
\sum_{\symCardinality=0}^\infty
\symRegProbability_
{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right]^{\symCount_\symRegVal}
\nonumber\\
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\cdot
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
\cdot
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}.
\end{align}
The final factorization shows that all register values are independent and identically distributed under the Poisson model. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange
\end{cases}
\end{equation}

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\vec{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}. Here $\vec{\symRegValVariate}$ denotes the vector of all observed register values. Each component of $\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The inverse Poisson transform gives
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Therefore, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$. Hence, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

\section{Original cardinality estimation method}
\label{sec:cardinality_estimation}
The original cardinality estimate \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal_\symIndexJ}$ distinct element insertions are needed until the value of register $\symIndexJ$ reaches $\symRegVal_\symIndexJ$. Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1},\ldots,\symNumReg2^{\symRegVal_\symNumReg}\rbrace$.

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityRawEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible compared to the estimation error.

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality. The chart is based on 10\,000 randomly generated HyperLogLog sketches. Obviously, the raw estimator is biased for small and large cardinalities and fails to return accurate estimates. Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.

@@

A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities. This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$. The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$ in case $\symRegRange \geq 30$. For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$, which would require some correction.

@@

Consider following integer distribution
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
This distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. The expectation of $2^{-\symRegVal}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegVal})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
 is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).

>>>>

\begin{document}
\maketitle
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle. The new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

Currently, the most memory efficient algorithm that also works for distributed setups is the near-optimal HyperLogLog algorithm \cite{Flajolet2007} with space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$. The originally proposed estimation method has some problems to guarantee the same estimation error over the entire range of cardinalities. It was proposed to correct the estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.

@@

\section{HyperLogLog data structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$. All registers start with initial value zero. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

@@

The sum $\symPrecision+\symRegRange$ corresponds to the number of consumed hash value bits and defines the maximum cardinality that can be tracked. Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent and the estimation accuracy is drastically reduced.

@@

\cref{alg:insert} has some properties which are especially useful in distributed environments. The insertion order of elements has no influence on the final HyperLogLog sketch. Furthermore, HyperLogLog sketches with same parameters $(\symPrecision, \symRegRange)$ can be easily merged. The resulting register values are simply given by the register-wise maximum values. The final HyperLogLog sketch corresponds to the union of both sets that have been represented by the two input sketches.

A $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}. In this case each register value is represented by a single bit. Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm for which $\symRegRange = 0$.

\begin{algorithm}
\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers. All register values $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) are initially equal to 0 and remain in the range $[0, \symRegRange+1]$.}
\label{alg:insert}
\begin{algorithmic}
\Procedure {InsertElement}{\symDataItem, $\vec{\symRegVal}$}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
\State $\symIndexJ \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
\State $\symRegVal' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
\State $\symRegVal_\symIndexJ \gets\max(\symRegVal_\symIndexJ, \symRegVal')$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

Under the assumption of a strong universal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
\begin{equation}
\symRegProbability_{\symNu\symKappa} 
:=
\begin{cases}
1 & \symNu=0 \wedge \symKappa = 0\\
0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
0 & \symNu\geq1 \wedge \symKappa = 0\\
\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
\end{cases}
\end{equation}

The order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$. As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation. The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim\vec{\symCount}-2$, respectively.

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Unfortunately, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze, because the register values are statistically dependent. Therefore, as proposed in \cite{Flajolet2007} a Poisson model can be used, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexJ=1}^\symNumReg
\symRegProbability_
{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
\nonumber\\
&= 
\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left[
\sum_{\symCardinality=0}^\infty
\symRegProbability_
{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right]^{\symCount_\symRegVal}
\nonumber\\
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\cdot
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
\cdot
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}.
\end{align}
The final factorization shows that all register values are independent and identically distributed under the Poisson model. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange
\end{cases}
\end{equation}

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\vec{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}. Here $\vec{\symRegValVariate}$ denotes the vector of all observed register values. Each component of $\vec{\symRegValVariate}$ is distributed according to \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The unique solution of this equation is given by
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Consequently, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$ which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

\section{Original cardinality estimation method}
\label{sec:cardinality_estimation}
The original cardinality estimator \cite{Flajolet2007} is based on the idea that approximately $\symNumReg  2^{\symRegVal-1}$ distinct element insertions are needed until the value of a register reaches $\symRegVal$. Given that, a rough cardinality estimate can be obtained by calculating the average over the values $\lbrace \symNumReg 2^{\symRegVal_1-1},\ldots,\symNumReg2^{\symRegVal_\symNumReg-1}\rbrace$.

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityRawEstimate
=
2\alpha_\symNumReg
\operatorname{HM}
\!\left(
\symNumReg 2^{\symRegVal_1-1},\ldots,\symNumReg2^{\symRegVal_\symNumReg-1}
\right)
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}.
\end{equation}
Here $2\alpha_\symNumReg$ is a bias correction factor. $\alpha_\symNumReg$ for a given number of registers $\symNumReg$ was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible compared to the estimation error.

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality. The chart is based on 10\,000 randomly generated HyperLogLog sketches. More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}. Obviously, the raw estimator is biased for small and large cardinalities and fails to return accurate estimates. Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.

@@

A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities. This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$. The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$ in case $\symRegRange \geq 30$. For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$ for which the raw estimator would return incorrect results.

@@

Consider following integer distribution
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. The expectation of $2^{-\symRegVal}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegVal})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
 is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).
